Multilevel computation of glacier geometry \\ from Stokes dynamics

### 2021-12-26 paper/msg.tex
Edits.

The standard model for determining the geometry of glaciers is based in part upon a dynamical description of ice flow, namely a shear-thinning version of the Stokes equations.  This Glen-Stokes ice-dynamics model can only determine the free-surface glacier geometry when it is combined with the surface kinematical equation (SKE) and a climate input, namely the balance of the rates of accumulation of snow and of ablation, i.e.~melting and runoff.  In this section we state the strong form of the resulting ``coupled'' glacier geometry model in the steady-state case, which we call the steady ice geometry problem (SIGP).  The emphasis here is on the complementarity-problem nature of the SIGP, which is often not explicit in the glaciers literature (e.g.~as observed by \cite{SchoofHewitt2013}).  We then define its weak form, noting along the way several unknown aspects of the theory.  Section \ref{sec:evolution} gives the straightfoward extension of the steady case to a time-discretized evolving-geometry model, the implicit ice geometry problem (IIGP).

The SIGP is defined on a fixed, bounded map-plane region $\Omega \subset \RR^d$ for $d=1,2$ (Figure \ref{fig:stokesdomain}).  The coordinates on $\Omega$ are denoted as $x$ when $d=1$ and $x,y$ when $d=2$.  On $\Omega$ we assume that a climatic mass balance (CMB) \cite{Cogleyetal2011} function $a(x,y)$ is defined at every point, whether or not ice is present at that location.  In ice-free areas this function is the ``potential'' CMB, namely the annual balance of snow accumulation and melt if ice were present, e.g.~as computed by energy balance in a climate or weather model.  Here we assume $a$ has units of ice thickness per time, compatible with our assumption of constant ice density (below).  Next we assume there is a bed elevation function $b(x,y)$ defined everywhere on $\Omega$, which determines the topography upon which the glacier sits.  The functions $a$ and $b$ are the data (inputs) of the SIGP.

@@

The Glen-Stokes dynamical model, the so-called stress balance, only applies in the icy domain in $\RR^{d+1}$.  We make a strong, but common \cite[for example]{IsaacStadlerGhattas2015,Jouvetetal2008,Lengetal2012,WirbelJarosch2020} assumption this this icy domain has a well-defined upper surface elevation, a function $s(x,y)$.  That is, we assume there are \emph{no overhangs}, and that the under-side of the ice is in contact with the bed $b$.  We then define $s$ everywhere in $\Omega$ by extending with $s=b$ where ice is absent, thus $s\ge b$ applies on all of $\Omega$.  Note that $s$ is part of the model solution; it is not given data.

However, as a consequence of the no-overhangs assumption, the SIGP as described here is likely not to be well-posed because of an issue at the ice margin.  There may be no steady state because the fluid in the vicinity of a steep ice margin, especially on a steep bed feature, ``wants'' to generate an overhang, violating the assumption that the function $s$ is well-defined.  Furthermore the same concern applies to each time step of an evolving model; overhangs can develop at the margin.  Because overhangs are small features in large glaciers and ice sheets, most modeling literature ignores this possibility and assumes well-defined surface elevation and thickness \cite{Jouvetetal2008,Lengetal2012,WirbelJarosch2020}.  Exceptions \cite{PralongFunk2005} and \cite{Jouvetetal2011} provide a continuum model in which ice-cliff calving occurs via a damage variable and a stress-fracture failure criterion, and such a model might explain how (essentially) well-defined surface elevation and thickness could arise in a well-posed model.  However, such a model is a nontrivial extension of the above equations, and though marginal cliffs may exist in our model (Section \ref{sec:fe}), overhangs are never allowed herein.

Based on a well-defined upper surface, the icy domain is the open set
\begin{equation}
\Lambda_s = \{(x,y,z)\,|\,(x,y) \in \Omega \,\text{ and }\, b(x,y) < z < s(x,y)\}  \subset \RR^{d+1}, \label{eq:lambdas}
\end{equation}
where $z$ is vertically-upward.  (Coordinates are denote $(x,z)$ on $\Lambda_s$ when $d=1$.)  The domain $\Lambda_s$ has the topology of the product of an open subset of $\Omega$ and an interval, and we will approximate it using an extruded mesh (Section \ref{sec:fe}).  Again note $\Lambda_s$ is part of the solution.

We model the ice as an incompressible, very-viscous \cite{Acheson1990}, non-Newtonian fluid subject to Glen's shear-thinning flow law \cite{GreveBlatter2009}; see also \cite[Chapter 1]{FowlerNg2021}.  Allowing any Glen exponent $\nn\ge 1$, the equations inside $\Lambda_s$ are
\begin{align}
- \nabla \cdot \tau + \nabla p &= \rhoi \bg &&\text{\emph{stress balance}} \label{eq:forcebalance} \\
\nabla \cdot \bu &= 0 &&\text{\emph{incompressibility}} \label{eq:incompressible} \\
\tau &= B_\nn |D\bu|^{(1/\nn) - 1} D\bu  &&\text{\emph{flow law}} \label{eq:viscflowlaw}
\end{align}
The solution fields here are the velocity $\bu$, pressure $p$, and deviatoric stress $\tau$.  Equations \eqref{eq:forcebalance} and \eqref{eq:viscflowlaw} form the momentum conservation (stress balance) model, while \eqref{eq:incompressible} follows from constant density and mass conservation.

Regarding tensors and their notation, recall that the (Cauchy) stress tensor $\sigma$ decomposes into the deviatoric part $\tau$ minus the pressure, i.e.~$\sigma = \tau - p\,I$, so equation \eqref{eq:forcebalance} simply says $-\Div \sigma = \rhoi \bg$.  The strain rate tensor $D\bu$ is the symmetric part of $\grad \bu$, $D\bu = \frac{1}{2} \left(\grad\bu + \grad\bu^\top\right)$, and the tensor norm used in \eqref{eq:viscflowlaw} satisfies $|D\bu|^2 = \frac{1}{2} (D\bu)_{ij} (D\bu)_{ij}$.  Because $D\bu$ is symmetric, and because it has trace zero by equation \eqref{eq:incompressible}, i.e.~$\trace(D\bu)=\nabla \cdot \bu = 0$, equation \eqref{eq:viscflowlaw} then implies that $\tau$ is also symmetric with trace zero, thus that $p=-(d+1)^{-1} \trace \sigma$.

@@

>>>>

The standard model for determining the geometry of glaciers is based upon a dynamical description of ice flow, namely a shear-thinning version of the Stokes equations.  However, this Glen-Stokes ice-dynamics model only determines the free-surface glacier geometry when combined with the surface kinematical equation (SKE) and a climate input, namely the balance of the rates of snow accumulation and melting/runoff (ablation).  In this section we state the strong form of the resulting ``coupled'' glacier geometry model in the steady-state case, which we call the steady ice geometry problem (SIGP).  The emphasis here is on the complementarity-problem nature of the SIGP, which is often not explicit in the glaciers literature (but see \cite{SchoofHewitt2013}).  We then define the weak form, noting along the way several unresolved aspects of the theory.  The steady model has a straightforward extension to a time-discretized evolving-geometry model, the implicit ice geometry problem (IIGP; Section \ref{sec:evolution}).

The SIGP is defined on a fixed, bounded map-plane region $\Omega \subset \RR^d$ for $d=1,2$ (Figure \ref{fig:stokesdomain}; coordinates on $\Omega$ are denoted as $x$ when $d=1$ and $x,y$ when $d=2$).  On $\Omega$ we assume that a time-independent climatic mass balance (CMB) \cite{Cogleyetal2011} function $a(x,y)$ is defined at every point, whether or not ice is present at that location.  In ice-free areas this function is the ``potential'' CMB, namely the annual balance of snow accumulation and melt if ice were present, e.g.~as computed by energy balance in a climate or weather model.  We assume $a$ has units of ice thickness per time, compatible with our assumption of constant ice density.  Also, a bed elevation function $b(x,y)$ is defined everywhere on $\Omega$; it gives the topography upon which the glacier sits.  The functions $a$ and $b$ are the data (inputs) of the SIGP.

@@

Now, the Glen-Stokes stress balance equations (below) only apply in a certain icy domain in $\RR^{d+1}$.  We make a strong, but common \cite[for example]{IsaacStadlerGhattas2015,Jouvetetal2008,Lengetal2012,WirbelJarosch2020} assumption that this domain has a well-defined upper surface elevation, a function $s(x,y)$.  That is, we assume there are \emph{no overhangs}, and that the under-side of the ice is in contact with the bed $b$.  We then define $s$ everywhere in $\Omega$ by extending with $s=b$ where ice is absent, thus $s\ge b$ applies on all of $\Omega$.  Note that $s$ is part of the model solution; it is not given data.

Based on this well-defined upper surface, the icy domain is the open set
\begin{equation}
\Lambda_s = \{(x,y,z)\,|\,(x,y) \in \Omega \,\text{ and }\, b(x,y) < z < s(x,y)\}  \subset \RR^{d+1}, \label{eq:lambdas}
\end{equation}
where $z$ is vertically-upward.  The domain $\Lambda_s$, which is a \emph{solution} in our model, has the topology of the product of an open subset of $\Omega$ and an interval, and we will approximate it using an extruded mesh (Section \ref{sec:fe}).

We model the ice as an incompressible, very-viscous \cite{Acheson1990}, non-Newtonian fluid in $\Lambda_s$.  Allowing any Glen exponent $\nn\ge 1$ \cite{GreveBlatter2009}, the equations are
\begin{align}
- \nabla \cdot \tau + \nabla p &= \rhoi \bg &&\text{\emph{stress balance}} \label{eq:forcebalance} \\
\nabla \cdot \bu &= 0 &&\text{\emph{incompressibility}} \label{eq:incompressible} \\
\tau &= B_\nn |D\bu|^{(1/\nn) - 1} D\bu  &&\text{\emph{flow law}} \label{eq:viscflowlaw}
\end{align}
The solution fields here are the velocity $\bu$, pressure $p$, and deviatoric stress $\tau$.  Equations \eqref{eq:forcebalance} and \eqref{eq:viscflowlaw} form the momentum conservation (stress balance) model, while \eqref{eq:incompressible} follows from constant density and mass conservation \cite[Chapter 1]{FowlerNg2021}.

Regarding tensors and their notation, recall that the Cauchy stress tensor $\sigma$ decomposes into the deviatoric part $\tau$ minus the pressure, i.e.~$\sigma = \tau - p\,I$, so equation \eqref{eq:forcebalance} simply says $-\Div \sigma = \rhoi \bg$.  The strain rate tensor $D\bu$ is the symmetric part of $\grad \bu$, $D\bu = \frac{1}{2} \left(\grad\bu + \grad\bu^\top\right)$, and the tensor norm used in \eqref{eq:viscflowlaw} satisfies $|D\bu|^2 = \frac{1}{2} (D\bu)_{ij} (D\bu)_{ij}$.  Because $D\bu$ is symmetric, and because it has trace zero by equation \eqref{eq:incompressible}, i.e.~$\trace(D\bu)=\nabla \cdot \bu = 0$, equation \eqref{eq:viscflowlaw} then implies that $\tau$ is also symmetric with trace zero, thus that $p=-\trace \sigma / (d+1)$.

@@

As a consequence of the no-overhangs assumption, the SIGP as described here is likely not to be well-posed because of an issue at the ice margin.  There may be no steady state because the fluid in the vicinity of a steep ice margin, especially on a steep bed feature, generates an overhang, violating the assumption that the function $s$ is well-defined.  Furthermore the same concern applies to each time step of an evolving model; overhangs can develop at the margin.  Because overhangs are small features in large glaciers and ice sheets, most modeling literature ignores this possibility and assumes well-defined surface elevation and thickness \cite{Jouvetetal2008,Lengetal2012,WirbelJarosch2020}.  The exceptional literature \cite{Jouvetetal2011,PralongFunk2005} provides a model in which ice-cliff calving occurs via a damage variable and a stress-fracture failure criterion, which might explain how a (nearly) well-defined surface elevation and thickness could arise in a more-complete, and well-posed, model.  However, such a model is a nontrivial extension of the above equations.  Neither marginal cliffs nor overhangs are allowed in our theory.
